---
title: "Data Wrangling with R"
author: "Abigail Sewell"
date: "2026-03-04"
format:
    html:
        toc: true
        toc-depth: 3
        toc-location: right-body
        embed-resources: true
execute:
    freeze: true
---
::: callout-note
**Note** This codebook will follow the structure and content of the published codebook for this project: https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r/
:::

# Overview

## Learning Outcomes

> By the end of this module, you will be able to:

-   Describe the concept of Data Wrangling.
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to “Tidy” data
-   Explain how to merge relational data sets using join functions. (Next module)
-   Explain how to use grouped mutates and filter together.
-   Be familiar with major dplyr functions for transforming data.
-   Create a new variable with mutate() and case_when().
-   Use the pipe operator to shape the data to prepare for analysis and visualization

## The testbook chapters cover
-   ch3: Data transformation
-   ch5: Data Tidying
-   ch13: Numbers

# Intro to Data Wrangling

## Loading the packages
-   to add a codechunk, use shortcut cmd+alt+i

```{r}
#options(repos = c(CRAN = "https://cloud.r-project.org"))
#install.packages(tidyverse)
#library(tidyverse)

```

# What is tidyverse
The tidyverse is a collection of r packges that share common design philosophy and are designed to work together seamlessly. the tidyverse includes packages for data manipulation, visualization, and modeling, among other tasks. some core packages are:
-   ggplot2
-   dplr
-   tidyr
-   readr
-   purrr
-   tibble
-   stringr
-   forcats


## Initial Data Preparation and Exploration

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))

install.packages("dplyr")
library(dplyr)

install.packages("nycflights13")
library(nycflights13)
```

# data wrangling with dplyr

Note: use shift, alt, i, for multi-cursor activision

airlines                      Airline names.
airports                      Airport metadata
flights                       Flights data
planes                        Plane metadata.
weather                       Hourly weather data


```{r}
library(tidyverse)
library(nycflights13)

flights |> 
  count(month) |> 
  mutate(month = as_factor(month),
         month = fct_reorder(month, n)) |>
  ggplot(aes(month, n)) +
  geom_col(fill = "#F3C4C7", color = "black") +
  coord_flip()+
  labs(x = "Month",
       y = "Number of Flights",
       title = "Number of flights by month during 2013 in New York")
```


